The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-03 VegOzone.tex
Some more text in Section 2. Started with section 3.

In our assessment on ozone damage on vegetation in the subarctic in relation to the 2018 heatwave, we focus on the NIBIO operated local site Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley (Section~\ref{subsec:svanhovd}). Throughout the paper, we will use Svanvik synonymously for the atmospheric monitoring site at Svanhovd. The available meteorological monitoring data are described in Section~\ref{subsubsec:atmo_svanvik}. In Section~\ref{subsubsec:ozone_garden}, the ozone garden -- a plant-based ozone monitoring -- will be described in more detail. For evaluation of the ozone data taken at Svanhovd, we compare with observational data from other monitoring sites in northern Fennoscandia and use as well as reanalysis products, i.a., as complement where observational data was not available or spatially sparse, (Section~\ref{subsec:obs_data}).

\subsection{NIBIO research station Svanhovd}
\label{subsec:svanhovd}
"Svanhovd is located beside the Pasvik river, in the middle of the wedge of Norwegian land separating Russia and Finland in the north. It is where the eastern Siberian taiga meets the western Boreal forest. [...] Svanhovd was originally established as a State Demonstration and Experimental Farm in the 1930's, in order to expand agriculture in Pasvik valley [...]" - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO webpage, accessed March 2020}.\\
The research station at Svanhovd comprises long-term observations of various agrameteorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil moisture and temperature at different depths.

For the purpose of relating observed damage on ozone sensitive plants (e.g. clover and tobacco) an ozone monitor had been installed during the growing seasons 2018/19. The relevant data are described in more detail Section~\ref{subsubsec:atmo_svanvik}.

@@

All agrameterological variables including temperature and precipitation are available from approximately 1992 to present day from LandbruksMeteorologisk Tjeneste provided by NIBIO (\href{https://lmt.nibio.no/}{lmt.nibio.no}, note the station name here is Pasvik).

{\bf TODO: (Technical) details about ozone monitoring at Svanhovd - historical period. Project related monitoring (growing seasons 2018/19.) $\rightarrow$ NILU -- Sverre?}

@@

\subsubsection{Ozone garden}
\label{subsubsec:ozone_garden}
      {\bf TODO: Details about the ozone garden.} \emph{Author: Ane/Aud -- pictures are welcome, just upload them on overleaf or send them.}

\subsection{Other data}
\label{subsec:obs_data}
To determine the significance of the conditions in 2018, we are interested not only in present (2018/19) surface ozone concentrations at Svanvik but also in long term observations and a climatology derived thereof. While data for the period 1986--1996 exists for Svanvik, no ozone measurements have been conducted at Svanhovd for the past $14\,\unit{years}$. Hence, we also have to have a look at ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are limited to a handful of atmospheric monitoring sites. An overview over their geographical locations is given in Fig.~\ref{fig:station_map_fennoscandia}. The data is descibed in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which was not available at all times or spatially sparse, we also look at verious ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking.
Note that from 1986--1996, ozone monitoring has been conducted operationally by NILU at Svanhovd.

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \label{fig:ozone_timesseries_fenoscandic_obs}
  \caption{Time series of ozone observations in northern Fennoscandia. Data taken from EBAS as well as from dedicated ozone monitoring at Svanhovd in the growing seasons 2018/19 (Tab.~\ref{tab:ebas_obs}). The hatched areas indicate time periods with insufficient quality control - \bf{TODO: citation NILU ozone report 200x}.}
\end{figure*}

@@

Different reanalysis products have been used to evaluate and complement the observational data.

There are two global reanalysis products of ECMWF available that include also ozone. In addition, higher resolution surface ozone reanalysis for Europe is available from Copernicus service for regional air quality control. The details are given in Tab.~\ref{tab:ozone_rea}. The covered time periods differ but are sufficient for our purpose of comparison of derived ozone climatologies.

@@

\section{Statistical analysis}
Statistical evaluation of Svanhovd (Svanvik/Pasvik) ozone data in a regional and climatological context.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{temperature_signific}
  \caption{Temperature at Svanvik. Percentage of days warmer/colder than $\pm 1\,\sigma$ from climatological mean displayed for each month. (a) 2018; (b) 2019. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/ right lower). The summer of 2018 was significantly warmer than average, especially May and July. The summer of 2019 was fairly average but significantly warmer than average in April and colder in July/August.}
  \label{fig:plot_temperature_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{precipitation_signific}
  \caption{Precipitation at Svanvik. Percentage of days wetter/drier than $\pm \frac{1}{2}\,\sigma$ from climatological mean displayed for each month. (a) 2018; (b) 2019. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/ right lower). The summer of 2018 was significantly drier than in 2019.}
  \label{fig:plot_precipitation_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_signific}
  \caption{Global irradiance at Svanvik. Percentage of days above/below $\pm 1\,\sigma$ from climatological mean displayed for each month. (a) 2018; (b) 2019. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/ right lower). In summer 2018 (May/July), gobal irradiance was significantly higher than on average, while 2019 was generally an average year.}
  \label{fig:global_rad_signific}
\end{figure}

>>>>

In our assessment on ozone damage on vegetation in the subarctic in relation to the 2018 heatwave, we focus on the NIBIO operated local site Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley (Section~\ref{subsec:svanhovd}). Throughout the paper, we will use Svanvik synonymously for the atmospheric monitoring site at Svanhovd. The available meteorological monitoring data are described in Section~\ref{subsubsec:atmo_svanvik}. In Section~\ref{subsubsec:ozone_garden}, the ozone garden -- a plant-based ozone monitoring -- will be described in more detail. Other data used for evaluation of the ozone data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use as well reanalysis products (Section~\ref{subsubsec:ozone_rea}) as complement when observational data was not available.

\subsection{NIBIO research station Svanhovd}
\label{subsec:svanhovd}
"Svanhovd is located beside the Pasvik river, in the middle of the wedge of Norwegian land separating Russia and Finland in the north. It is where the eastern Siberian taiga meets the western Boreal forest. [...] Svanhovd was originally established as a State Demonstration and Experimental Farm in the 1930's, in order to expand agriculture in Pasvik valley [...]" - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO webpage, accessed March 2020}.\\
The research station at Svanhovd comprises long-term observations of various agrameteorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

For the purpose of relating observed damage on ozone sensitive plants (e.g. clover and tobacco) an ozone monitor had been installed during the growing seasons 2018/19. The relevant data are described in more detail in Section~\ref{subsubsec:atmo_svanvik}.

@@

All agrameterological variables including temperature and precipitation are available from September 1992 to present day (LandbruksMeteorologisk Tjeneste provided by NIBIO -- \href{https://lmt.nibio.no/}{lmt.nibio.no}, note the station name here is Pasvik).

{\bf TODO: Some more (technical) details about ozone and atmospheric monitoring at Svanhovd. $\rightarrow$ Sverre?}

@@

\subsubsection{Ozone garden}
\label{subsubsec:ozone_garden}
      {\bf TODO: Details about the ozone garden and observed damage on ozone sensitive plants. We will need to come back to this in the discussion section after the Do3se modelling part.} \emph{Author: Ane/Aud -- pictures are welcome, just upload them on overleaf or send them to me.}

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we have to derive a surface ozone concentration climatology at Svanvik. From 1986--1996 ozone monitoring has been operated by NILU at Svanhovd, but no ozone measurements have been conducted for the past $14\,\unit{years}$. Due to probabel changes in tropospheric background ozone and a reduction of summertime peak values related to precursor that fell under airquality regulations, we also have to have a look at ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted atmospheric monitoring in past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data is descibed in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which is not available at all times and spatially sparse, we also look at verious ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observation are locate on higher ground than Svanhovd which will result in higher surface ozone concentrations compared to data taken at Svanvik as ozone abundance decreases with altitude in the absence of local precursors. {\bf TODO: The following has to be confirmed and ellaborated on!} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

Data from before the 1990s is available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care {\bf TODO: citation NILU ozone report 200x}. Especially, ozone monitors did not undergo a regular re-calibration. This leads likely to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul had been relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have intruduced a shift in the data series denotes as Jergul/Karasjok in the following. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlapp of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas herin indicate the time periods with insufficient quality control as mentioned above.

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}). Data taken from EBAS. The hatched areas indicate time periods with insufficient quality control. \bf{TODO: citation NILU ozone report 200x}.}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

Details of renalysis data sets refered to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone (MACC and CAMSRA). The temporal as well as spatial resolution of these reanalysis products is rather corse: 3-hourly and $0.75^\circ\times 0.75^\circ$ or roughly $(78.1\times 83.4)\,\unit{km}$ at the location of Svanvik, respectively. In addition, higher resolution surface ozone reanalysis for whole Europe is available from Copernicus service for regional air quality control. The regional model reanalysis ensemble mean is at fairly high spatial and temporal resolution compared to the data from ECMWF: $0.1^\circ\times 0.1^\circ$ or roughly $(10.4\times 11.1)\,\unit{km}$ and 1-hourly, respectively. The covered time periods differ and no data are available for before the turn of the millenium. The CAMSRA and CAMS Regional Air Quality ensemble mean are available in near realtime, but only former covers a period of sufficient length for climate analysis (2003 -- present). Their quality with respect to observations will be discussed in Section~\ref{subsubsec:clim_ozone}. Without too much foreshadowing, the results prevented us from fetching the whole CAMSRA data.

@@

\section{Statistical analysis of atmospheric key variables}
In this section, we will evaluate atmospheric conditions in the growing seasons of 2018/19 by statistical means. We will derive climatologies for both, temperature, precipitation, global irradiance, and ozone for Svanvik (Section~\ref{subsec:climatologies}). The ozone measurements will be put into a climatological as well as regional context. We will look at anomalies of the atmospheric key variable mentioned above (Section~\ref{subsec:anomalies}).

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{temperature_signific}
  \caption{Temperature at Svanvik. Percentage of days warmer/colder than $\pm 1\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. The summer of 2018 was significantly warmer than average, especially May and July. The summer of 2019 was fairly average but significantly warmer than average in April and colder in July/August.}
  \label{fig:plot_temperature_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{precipitation_signific}
  \caption{Precipitation at Svanvik. Percentage of days wetter/drier than $\pm \frac{1}{2}\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above $\pm\frac{1}{2}\,\sigma$ if a normal distribution is assumed ($30.8\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. The summer of 2018 was significantly drier than in 2019.}
  \label{fig:plot_precipitation_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_signific}
  \caption{Global irradiance at Svanvik. Percentage of days above/below $\pm 1\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. In summer 2018 (May/July), gobal irradiance was significantly higher than on average, while 2019 was generally an average year.}
  \label{fig:global_rad_signific}
\end{figure}
